---
title: "Making hyperspectral figures worth looking at"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Making hyperspectral figures worth looking at}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 4.2, dpi = 110, fig.align = "center"
)
library(hyperspectaculR)
```

A hyperspectral cube holds a full spectrum at every pixel, and almost every
figure made from one throws that away — a single band, or three bands pushed
into red, green and blue. `hyperspectaculR` is for the other kind of figure:
compositions that could not be produced from any single wavelength, made to be
looked at rather than measured off.

It reads nothing and analyses nothing. Reading belongs to
[`tivis.r`](https://github.com/CTTIR/tivis.r) and
[`cuvis.r`](https://github.com/CTTIR/cuvis.r); the cube class and the analysis
pipeline belong to [`hyperspectR`](https://github.com/CTTIR/hyperspectR). This
package takes an `hsi_cube` and returns a `ggplot`.

Everything below runs on a small synthetic cube bundled with the package, so
the vignette builds without any recorded data.

```{r}
cube <- hsa_demo_cube()
dim(cube$data)
range(cube$wavelengths)
```

## Three compositions

### A radial spectral sweep

`hsa_mandala()` draws concentric annuli, each taken from a different
wavelength: the innermost ring is the shortest wavelength, the outermost the
longest. One image, sweeping the whole spectral axis outward from a centre.

```{r mandala}
hsa_mandala(cube)
```

Fewer rings give bolder banding; more rings sample the spectrum finely enough
that the transition reads as continuous.

```{r mandala-rings, fig.height = 3.2}
hsa_mandala(cube, n_rings = 8)
```

The centre is yours to choose, which matters when the subject is not in the
middle of the frame.

```{r mandala-centre, fig.height = 3.2}
hsa_mandala(cube, centre = c(20, 15), n_rings = 24)
```

### How much a pixel moves across the spectrum

`hsa_spectral_flux()` sums the absolute change between consecutive bands. Flat
spectra go dark; pixels whose reflectance swings across the spectrum light up.

```{r flux}
hsa_spectral_flux(cube)
```

This is not a variance map. Variance is blind to the order of the bands,
whereas this responds to spectral *shape* — two pixels with identical variance
score differently if one changes smoothly and the other oscillates.

### Colour from band groups

`hsa_fusion()` builds a composite by averaging three groups of bands rather
than picking three wavelengths. Averaging suppresses per-band sensor noise, so
the result is smoother and more saturated than a three-band composite, while
remaining a simple and disclosable operation.

```{r fusion}
hsa_fusion(cube)
```

Groups can be given as wavelengths when that is how you think about the
subject:

```{r fusion-wl}
wl <- cube$wavelengths
hsa_fusion(cube, red = wl[22], green = wl[12], blue = wl[2], by = "wavelength")
```

## Saying what you did to the image

Contrast stretching is what makes hyperspectral data legible on a screen. It is
also the easiest way to make a figure imply something the data does not. So
every function here captions the stretch it applied and the limits it used:

```{r caption}
hsa_mandala(cube)$labels$caption
hsa_spectral_flux(cube, stretch = "range")$labels$caption
hsa_mandala(cube, stretch = "none")$labels$caption
```

That caption travels with the figure. If a reviewer asks what was done to the
image, the answer is printed underneath it.

## Colour that survives the journey

`hsa_palette()` offers only ramps that are monotonic in lightness, so ordering
is preserved in greyscale printing and remains legible to colourblind readers.

```{r palette}
head(hsa_palette("magma", n = 6))
```

Rainbow-like ramps are excluded on purpose, with one exception kept for
screen-only use — and it tells you:

```{r turbo, warning = TRUE}
invisible(hsa_palette("turbo", n = 4))
```

## Composing figures

Everything is a `ggplot`, so panels combine with `patchwork`, re-theme, and
export through `ggsave()` at whatever size and resolution a journal wants.
Nothing writes a file as a side effect.

```{r patchwork, eval = requireNamespace("patchwork", quietly = TRUE), fig.width = 7, fig.height = 3}
library(patchwork)
hsa_fusion(cube) + hsa_spectral_flux(cube)
```

For print, export as vector where the composition allows it, or at 300–600 dpi
where it does not — these are raster images, so a high-resolution PNG or TIFF
is usually the honest choice:

```{r save, eval = FALSE}
ggplot2::ggsave("figure-1.png", hsa_mandala(cube),
                width = 180, height = 130, units = "mm", dpi = 600)
```

## Working from real recordings

With `hyperspectR` installed, the same functions take real cubes from either
supported instrument:

```{r real, eval = FALSE}
library(hyperspectR)

# TIVITA, via tivis.r -- no vendor SDK needed
cube <- hs_read_tivita("2019_11_25_13_29_24_SpecCube.dat")

# Cubert, via cuvis.r and the CUVIS SDK
# cube <- hs_read_cubert("session.cu3s")

hsa_fusion(cube, red = 850, green = 650, blue = 550, by = "wavelength")
```

Both readers return the same `hsi_cube`, so nothing here changes between
instruments — only the wavelengths in the caption.
