---
title: "Making hyperspectral figures worth looking at"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Making hyperspectral figures worth looking at}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 4.2, dpi = 110, fig.align = "center"
)
library(hyperspectaculR)
```

A hyperspectral cube holds a full spectrum at every pixel, and almost every
figure made from one throws that away — a single band, or three bands pushed
into red, green and blue. `hyperspectaculR` is for the other kind of figure:
compositions that could not be produced from any single wavelength, made to be
looked at rather than measured off.

It reads nothing and analyses nothing. Reading belongs to
[`tivis.r`](https://github.com/CTTIR/tivis.r) and
[`cuvis.r`](https://github.com/CTTIR/cuvis.r); the cube class and the analysis
pipeline belong to [`hyperspectR`](https://github.com/CTTIR/hyperspectR). This
package takes an `hsi_cube` and returns a `ggplot`.

Everything below runs on a small synthetic cube bundled with the package, so
the vignette builds without any recorded data.

```{r}
cube <- hsa_demo_cube()
dim(cube$data)
range(cube$wavelengths)
```

## Start with the distribution

Before any of the images, it is worth looking at what is actually in the cube.
`hsa_spectral_density()` bins every pixel's reflectance against wavelength and
draws the joint distribution, with the mean spectrum and a 5–95% envelope
threaded through it.

```{r density}
hsa_spectral_density(cube)
```

This is the only composition here that is not an image, and it is the one the
others should be read against. Every image below applies a contrast stretch;
this shows the distribution that stretch is applied to. Passing
`show_limits = TRUE` draws those limits straight onto it, so you can see
whether they sit sensibly inside the data or are cutting into it.

```{r density-limits, fig.height = 3.6}
hsa_spectral_density(cube, show_limits = TRUE)
```

Two details matter for trusting the figure. Values outside the range are
**dropped, not clipped** — clipping would pile their mass into the end bins and
manufacture bright edges at the extremes, which is exactly the artefact this
plot exists to expose. And counts are shown on a `log1p` scale by default,
because reflectance histograms are heavy-tailed and a linear scale shows only
the mode; the colourbar labels are back-transformed, so the legend still reads
true counts.

On real recordings this is often where problems surface first. A capture whose
mass piles onto a few discrete reflectance values is quantised — too few
photons divided by a white reference — and no amount of careful rendering
downstream will put information back.

## Four compositions

### A radial spectral sweep

`hsa_mandala()` draws concentric annuli, each taken from a different
wavelength: the innermost ring is the shortest wavelength, the outermost the
longest. One image, sweeping the whole spectral axis outward from a centre.

```{r mandala}
hsa_mandala(cube)
```

Fewer rings give bolder banding; more rings sample the spectrum finely enough
that the transition reads as continuous.

```{r mandala-rings, fig.height = 3.2}
hsa_mandala(cube, n_rings = 8)
```

The centre is yours to choose, which matters when the subject is not in the
middle of the frame.

```{r mandala-centre, fig.height = 3.2}
hsa_mandala(cube, centre = c(20, 15), n_rings = 24)
```

### How much a pixel moves across the spectrum

`hsa_spectral_flux()` sums the absolute change between consecutive bands. Flat
spectra go dark; pixels whose reflectance swings across the spectrum light up.

```{r flux}
hsa_spectral_flux(cube)
```

This is not a variance map. Variance is blind to the order of the bands,
whereas this responds to spectral *shape* — two pixels with identical variance
score differently if one changes smoothly and the other oscillates.

### Colour from band groups

`hsa_fusion()` builds a composite by averaging three groups of bands rather
than picking three wavelengths. Averaging suppresses per-band sensor noise, so
the result is smoother and more saturated than a three-band composite, while
remaining a simple and disclosable operation.

```{r fusion}
hsa_fusion(cube)
```

Groups can be given as wavelengths when that is how you think about the
subject:

```{r fusion-wl}
wl <- cube$wavelengths
hsa_fusion(cube, red = wl[22], green = wl[12], blue = wl[2], by = "wavelength")
```

## Saying what you did to the image

The density plot above shows the distribution; the images apply a stretch to
it. That stretch is what makes hyperspectral data legible on a screen, and also
the easiest way to make a figure imply something the data does not. So every
function here captions the stretch it applied and the limits it used:

```{r caption}
hsa_mandala(cube)$labels$caption
hsa_spectral_flux(cube, stretch = "range")$labels$caption
hsa_mandala(cube, stretch = "none")$labels$caption
hsa_spectral_density(cube)$labels$caption
```

That caption travels with the figure. If a reviewer asks what was done to the
image, the answer is printed underneath it.

## Colour that survives the journey

`hsa_palette()` offers only ramps that are monotonic in lightness, so ordering
is preserved in greyscale printing and remains legible to colourblind readers.

```{r palette}
head(hsa_palette("magma", n = 6))
```

Rainbow-like ramps are excluded on purpose, with one exception kept for
screen-only use — and it tells you:

```{r turbo, warning = TRUE}
invisible(hsa_palette("turbo", n = 4))
```

## Composing figures

Everything is a `ggplot`, so panels combine with `patchwork`, re-theme, and
export through `ggsave()` at whatever size and resolution a journal wants.
Nothing writes a file as a side effect.

```{r patchwork, eval = requireNamespace("patchwork", quietly = TRUE), fig.width = 7, fig.height = 3}
library(patchwork)
hsa_fusion(cube) + hsa_spectral_flux(cube)
```

For print, export as vector where the composition allows it, or at 300–600 dpi
where it does not — these are raster images, so a high-resolution PNG or TIFF
is usually the honest choice:

```{r save, eval = FALSE}
ggplot2::ggsave("figure-1.png", hsa_mandala(cube),
                width = 180, height = 130, units = "mm", dpi = 600)
```

## Working from real recordings

With `hyperspectR` installed, the same functions take real cubes from either
supported instrument:

```{r real, eval = FALSE}
library(hyperspectR)

# TIVITA, via tivis.r -- no vendor SDK needed
cube <- hs_read_tivita("2019_11_25_13_29_24_SpecCube.dat")

# Cubert, via cuvis.r and the CUVIS SDK
# cube <- hs_read_cubert("session.cu3s")

hsa_fusion(cube, red = 850, green = 650, blue = 550, by = "wavelength")
```

Both readers return the same `hsi_cube`, so nothing here changes between
instruments — only the wavelengths in the caption.

A reasonable habit with an unfamiliar recording is to run
`hsa_spectral_density(cube, show_limits = TRUE)` first. If the distribution is
banded, truncated, or piled against one end, that is worth knowing before
choosing how to render it — and worth saying in the figure caption either way.
